---
title: "Project report"
author: "Vaughn Hendrix"
date: "2025-07-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

------------------------------------------------------------------------

## Introduction

The goal of this project is to explore the relationship between El Nino Southern Oscillation (ENSO) and tornado formation in the United States. ENSO is a global weather phenomenon located in the tropical pacific. It is a recurring pattern in which warmer sea surface temperatures and lower pressure shift between the east and west tropical pacific which explains the "oscillation" in its name. Sea surface temperature anomalies are explained in three phases: Neutral which is the average temperature over a given number of years, El Niño phase which is warmer temperatures than average in the eastern tropical pacific, while La Nina is warmer than average temperatures in the western tropical pacific and colder than average temperatures in the east tropical pacific. Consequently, ENSO may provide the necessary conditions, such as moist air, strong updrafts, and wind shear (the change in wind vector with height), contributing to tornado formation.

The aim of this analysis is to determine if the phase of ENSO plays a role in determining the frequency of tornadoes through the use of various questions:

-   Is there a trend in the appearance of tornadoes over the years? Does the phase of ENSO play a role in how many tornadoes are forming?

-   Does the phase of ENSO shift the locations of tornadoes?

-   How are the tornadoes themselves affected? Does their strength and size dependent on the phase of ENSO?

------------------------------------------------------------------------

## Data

#### Data Sources

My first data set comes from the NOAA Physical Sciences Laboratory: ERSSTv5 Niño 3.4 Index, available at [https://psl.noaa.gov/data/timeseries/month/](#0). The data set contains two variables: the observation date and the corresponding ENSO value. It provides monthly observations of sea surface temperature (sst) anomalies from 1948-present, totaling 936 entries.

My tornado data comes from the National Centers for Environmental Information (NCEI), available at <https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/>. The site provides CSV files containing storm details, fatalities, and location information from 1950-present. I used the storm details data sets, which include variables such as magnitude, intensity, event type, date, and more. Each data set contains 51 variables and between 55,000 to 80,000 observations, depending on the year.

To explore the relationship between ENSO and tornado formation, I first chose to use the nino3.4 index because it was a more general data set covering characteristics in both the Central and Eastern Pacific compared to others. Next, I analyzed 2007-2024 to see changes over time and make the data set more manageable. Additionally, the Enhanced Fujita (EF) scale was introduced in 2007 and limiting the time frame to begin at 2007 keeps the tornado scaling consistent.

#### Data Cleaning

First, I imported the data through the CSV covering storm details from 2007 to 2024.

```{r}
nino3.4 <- read_csv("nina34.anom.csv", show_col_types = FALSE)

tornadoes_2007 <- read_csv("StormDetails_2007.csv.gz", show_col_types = FALSE)
tornadoes_2008 <- read_csv("StormDetails_2008.csv.gz", show_col_types = FALSE)
tornadoes_2009 <- read_csv("StormDetails_2009.csv.gz", show_col_types = FALSE)
tornadoes_2010 <- read_csv("StormDetails_2010.csv.gz", show_col_types = FALSE)
tornadoes_2011 <- read_csv("StormDetails_2011.csv.gz", show_col_types = FALSE)
tornadoes_2012 <- read_csv("StormDetails_2012.csv.gz", show_col_types = FALSE)
tornadoes_2013 <- read_csv("StormDetails_2013.csv.gz", show_col_types = FALSE)
tornadoes_2014 <- read_csv("StormDetails_2014.csv.gz", show_col_types = FALSE)
tornadoes_2015 <- read_csv("StormDetails_2015.csv.gz", show_col_types = FALSE)
tornadoes_2016 <- read_csv("StormDetails_2016.csv.gz", show_col_types = FALSE)
tornadoes_2017 <- read_csv("StormDetails_2017.csv.gz", show_col_types = FALSE)
tornadoes_2018 <- read_csv("StormDetails_2018.csv.gz", show_col_types = FALSE)
tornadoes_2019 <- read_csv("StormDetails_2019.csv.gz", show_col_types = FALSE)
tornadoes_2020 <- read_csv("StormDetails_2020.csv.gz", show_col_types = FALSE)
tornadoes_2021 <- read_csv("StormDetails_2021.csv.gz", show_col_types = FALSE)
tornadoes_2022 <- read_csv("StormDetails_2022.csv.gz", show_col_types = FALSE)
tornadoes_2023 <- read_csv("StormDetails_2023.csv.gz", show_col_types = FALSE)
tornadoes_2024 <- read_csv("StormDetails_2024.csv.gz", show_col_types = FALSE)
#removed rows 31,35,36,37,39
```

Then, I changed one of the column names to enhance readability. I filtered out ENSO values that were either `NA` or `-9999`, as both represent invalid data. I then created a `BEGIN_YEARMONTH` column for joining it with the main data set. A `YEAR` column was created to filter for the years 2007 to 2024, and a `PHASE` column was added to label each observation as either El Niño (value \> 0.5), La Nina (value \< -0.5), or Neutral (−0.5 ≤ value ≤ 0.5).

```{r}
colnames(nino3.4)[2] <- "nino3.4"

nino3.4 <- nino3.4 |> 
  filter(!is.na(Date), 
         !is.na(nino3.4),
         nino3.4 != -9999) |>
  mutate(BEGIN_YEARMONTH = as.numeric(format(Date, "%Y%m")), 
         YEAR = year(Date),
         PHASE = case_when(
           nino3.4 > 0.5 ~ "El Nino", 
           nino3.4 >= -0.5 & nino3.4 <= 0.5 ~ "Neutral", 
           nino3.4 < -0.5 ~ "La Nina", 
           TRUE ~ "Other" )) |>
  filter(YEAR >= 2007, YEAR < 2025)
nino3.4$PHASE <- factor(nino3.4$PHASE)
```

To simplify the process, I created a few functions to streamline the workflow across all data sets. The first function filters the data to include only tornado observations. The second function selects the relevant variables from each data set. The final function uses a dictionary to map the old tornado rating scale to the modern EF scale. Since the data set contained both the older and modern rating system, this function ensures consistency by converting the old ratings to the modern scale.

```{r}
tornado_obs <- function(data) {
  data <- data |> filter(EVENT_TYPE == "Tornado")
  return(data)
}

filtering_data <- function(data) {
  data <- data |> 
    select(BEGIN_YEARMONTH, BEGIN_DAY, STATE, YEAR, MONTH_NAME, EVENT_TYPE, BEGIN_DATE_TIME, END_DATE_TIME, DAMAGE_PROPERTY, TOR_F_SCALE, TOR_LENGTH, TOR_WIDTH, BEGIN_LAT, BEGIN_LON)
   return(data)
}

F_SCALE <- c("F0" = "EF0", "F1" = "EF1", "F2" = "EF2", "F3" = "EF3", "F4" = "EF4", "F5" = "EF5")

F_SCALE_EQUALIZATION <- function(data) {
  data <- data |>
    mutate(TOR_F_SCALE = ifelse(TOR_F_SCALE %in% names(F_SCALE), F_SCALE[TOR_F_SCALE] , TOR_F_SCALE))
   return(data)
}
```

```{r}
tornadoes_2007 <- tornado_obs(tornadoes_2007)
tornadoes_2008 <- tornado_obs(tornadoes_2008)
tornadoes_2009 <- tornado_obs(tornadoes_2009)
tornadoes_2010 <- tornado_obs(tornadoes_2010)
tornadoes_2011 <- tornado_obs(tornadoes_2011)
tornadoes_2012 <- tornado_obs(tornadoes_2012)
tornadoes_2013 <- tornado_obs(tornadoes_2013)
tornadoes_2014 <- tornado_obs(tornadoes_2014)
tornadoes_2015 <- tornado_obs(tornadoes_2015)
tornadoes_2016 <- tornado_obs(tornadoes_2016)
tornadoes_2017 <- tornado_obs(tornadoes_2017)
tornadoes_2018 <- tornado_obs(tornadoes_2018)
tornadoes_2019 <- tornado_obs(tornadoes_2019)
tornadoes_2020 <- tornado_obs(tornadoes_2020)
tornadoes_2021 <- tornado_obs(tornadoes_2021)
tornadoes_2022 <- tornado_obs(tornadoes_2022)
tornadoes_2023 <- tornado_obs(tornadoes_2023)
tornadoes_2024 <- tornado_obs(tornadoes_2024)
```

```{r}
tornadoes_2007 <- filtering_data(tornadoes_2007)
tornadoes_2008 <- filtering_data(tornadoes_2008)
tornadoes_2009 <- filtering_data(tornadoes_2009)
tornadoes_2010 <- filtering_data(tornadoes_2010)
tornadoes_2011 <- filtering_data(tornadoes_2011)
tornadoes_2012 <- filtering_data(tornadoes_2012)
tornadoes_2013 <- filtering_data(tornadoes_2013)
tornadoes_2014 <- filtering_data(tornadoes_2014)
tornadoes_2015 <- filtering_data(tornadoes_2015)
tornadoes_2016 <- filtering_data(tornadoes_2016)
tornadoes_2017 <- filtering_data(tornadoes_2017)
tornadoes_2018 <- filtering_data(tornadoes_2018)
tornadoes_2019 <- filtering_data(tornadoes_2019)
tornadoes_2020 <- filtering_data(tornadoes_2020)
tornadoes_2021 <- filtering_data(tornadoes_2021)
tornadoes_2022 <- filtering_data(tornadoes_2022)
tornadoes_2023 <- filtering_data(tornadoes_2023)
tornadoes_2024 <- filtering_data(tornadoes_2024)
```

```{r}
tornadoes_2007 <- F_SCALE_EQUALIZATION(tornadoes_2007)
tornadoes_2008 <- F_SCALE_EQUALIZATION(tornadoes_2008)
tornadoes_2009 <- F_SCALE_EQUALIZATION(tornadoes_2009)
tornadoes_2010 <- F_SCALE_EQUALIZATION(tornadoes_2010)
tornadoes_2011 <- F_SCALE_EQUALIZATION(tornadoes_2011)
tornadoes_2012 <- F_SCALE_EQUALIZATION(tornadoes_2012)
tornadoes_2013 <- F_SCALE_EQUALIZATION(tornadoes_2013)
tornadoes_2014 <- F_SCALE_EQUALIZATION(tornadoes_2014)
tornadoes_2015 <- F_SCALE_EQUALIZATION(tornadoes_2015)
tornadoes_2016 <- F_SCALE_EQUALIZATION(tornadoes_2016)
tornadoes_2017 <- F_SCALE_EQUALIZATION(tornadoes_2017)
tornadoes_2018 <- F_SCALE_EQUALIZATION(tornadoes_2018)
tornadoes_2019 <- F_SCALE_EQUALIZATION(tornadoes_2019)
tornadoes_2020 <- F_SCALE_EQUALIZATION(tornadoes_2020)
tornadoes_2021 <- F_SCALE_EQUALIZATION(tornadoes_2021)
tornadoes_2022 <- F_SCALE_EQUALIZATION(tornadoes_2022)
tornadoes_2023 <- F_SCALE_EQUALIZATION(tornadoes_2023)
tornadoes_2024 <- F_SCALE_EQUALIZATION(tornadoes_2024)
```

To create the main data set, I used the `bind_rows()` function to merge all tornado data frames from 2007 to 2024. I then converted the `MONTH_NAME` column into a factor with levels ordered chronologically to ensure the months appeared in the correct order.

```{r}
masterData <- bind_rows(tornadoes_2007, tornadoes_2008, tornadoes_2009, tornadoes_2010, tornadoes_2011, tornadoes_2012, tornadoes_2013, tornadoes_2014, tornadoes_2015, tornadoes_2016, tornadoes_2017, tornadoes_2018, tornadoes_2019, tornadoes_2020, tornadoes_2021, tornadoes_2022, tornadoes_2023, tornadoes_2024)

masterData <- masterData |>
  mutate(MONTH_NAME = factor(MONTH_NAME, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")))
```

Finally, I used the `left_join()` function to merge the `masterData` data set with the `nino3.4` data set, ensuring each tornado observation included the corresponding ENSO value and phase.

```{r}
masterData <- masterData |> left_join(nino3.4 |> select(nino3.4, BEGIN_YEARMONTH, PHASE), by = "BEGIN_YEARMONTH")
```

------------------------------------------------------------------------

## Results

#### Is there a trend in the appearance of tornadoes over the years? Does the phase of ENSO play a role in how many tornadoes are forming?

```{r}
trend <- masterData |> group_by(YEAR) |> summarise(count = n()) |> ungroup()
trend |> ggplot(aes(x = YEAR, y = count)) +
  geom_line() +
  scale_x_continuous(breaks = unique(masterData$YEAR), minor_breaks = NULL)
```

This time series shows the number of tornadoes since 2007. This graph shows a semi-periodic pattern, where it is common for a high to be followed immediately by a low, with a standard distance being three years. Recently, the trend for tornado appearance is on the rise, but from the semi-periodic distribution, there could be a decline.

```{r}
masterData |> 
  ggplot(aes(x = BEGIN_DAY)) + 
  geom_bar() +
  facet_wrap(~MONTH_NAME)
```

This graph displays the daily tornado count throughout the year. It shows that tornado activity typically peaks between April and June. In addition to the semi-periodic trend observed across multiple years, tornado activity within a single year also follows a seasonal pattern in spring and early summer tornado spikes, then declines in the latter half of the year. This seasonal peak could result from a southward dip in the jet streak, providing strong wind shear along with increasing daylight exposure and surface heating, creating good environments of instability.

```{r}
masterData |> 
  filter(PHASE == "El Nino") |>
  ggplot(aes(x = BEGIN_DAY, fill = PHASE)) + 
  geom_bar() +
  facet_wrap(~MONTH_NAME) +
  scale_fill_manual(values = c("El Nino" = "red"))
```

This graph displays the daily tornado count throughout the year, limited to the El Nino phase, showing that tornado activity peaks in April and May, with relatively little activity before or after. El Nino years exhibit a noticeably shorter and more concentrated peak period compared to the overall seasonal tornado peak from April to June.

```{r}
masterData |> 
  filter(PHASE == "La Nina") |>
  ggplot(aes(x = BEGIN_DAY, fill = PHASE)) + 
  geom_bar() +
  facet_wrap(~MONTH_NAME) +
  scale_fill_manual(values = c("La Nina" = "blue"))
```

This graph displays the daily tornado count throughout the year, limited to the La Nina phase, showing that tornado activity shows an extended peak from March through June, with relatively little activity before or after. Compared to the overall tornado season, La Nina years exhibit a longer and more active tornado period.

```{r}
masterData |> 
  filter(PHASE == "Neutral") |>
  ggplot(aes(x = BEGIN_DAY, fill = PHASE)) + 
  geom_bar() +
  facet_wrap(~MONTH_NAME) +
  scale_fill_manual(values = c("Neutral" = "grey"))
```

This graph displays the daily tornado count throughout the year, limited to the Neutral phase. Similar to La Nina years, tornado activity during the Neutral phase extends from April through July, indicating a longer and more sustained tornado season.

#### Does the phase of ENSO shift the locations of tornadoes?

```{r}
states <- map_data("state")
byState <- masterData |> group_by(STATE, PHASE) |> summarise(count = n(), .groups = "drop")
byState <- byState |> mutate(region = tolower(STATE)) |> select(-STATE)
tornado_map <- byState |> left_join(states, by = "region", relationship = "many-to-many")
```

```{r, fig.width = 20, fig.height = 5}
tornado_map |>
  ggplot(aes(x = long, y = lat, fill = count)) +
  geom_polygon(aes(group = group)) +
  scale_fill_viridis_c(option = "viridis", na.value = "gray90") +
  labs(title = "Tornado Concentration by Phase") +
  facet_wrap(~PHASE)
```

The following choropleth map of the United States shows the distribution of tornado activity, where the shade of the state represents the number of tornado observations. Across all three maps, it is common for tornadoes to be observed in the Southern and Central Midwest, but the range of activity varies depending on the phase. This specific activity region could be attributed to the warm and moist air from the Gulf of Mexico interacting with the cold and dry air of Canada which contribute key variables in tornado formation.

It can be seen that the range of activity of El Nino shrinks in comparison to the Neutral phase, being more limited towards the Southern region. This is likely due to the jet stream, where after being shifted southward by El Nino, creates a wetter and colder environment in the South, thus a more stable environment. Unlike El Nino, La Nina range of activity expands, covering more of the Midwestern region relative to the number of observation in comparison to the Neutral phase. With the jet stream moving polewards during La Nina, a larger area is warmer and wetter, creating a more favorable condition for tornadoes.

#### How are the tornadoes themselves affected? Does their strength and size dependent on the phase of ENSO?

```{r}
masterData |> mutate(DAMAGE_PROPERTY = substr(DAMAGE_PROPERTY, nchar(DAMAGE_PROPERTY), nchar(DAMAGE_PROPERTY))
) |> distinct(DAMAGE_PROPERTY)

masterData <- masterData |>
  mutate(DAMAGE_PROPERTY = case_when(
  substr(DAMAGE_PROPERTY, nchar(DAMAGE_PROPERTY), nchar(DAMAGE_PROPERTY)) == "K" ~ parse_number(DAMAGE_PROPERTY) * 10^3,
  substr(DAMAGE_PROPERTY, nchar(DAMAGE_PROPERTY), nchar(DAMAGE_PROPERTY)) == "M" ~ parse_number(DAMAGE_PROPERTY) * 10^6,
  substr(DAMAGE_PROPERTY, nchar(DAMAGE_PROPERTY), nchar(DAMAGE_PROPERTY)) == "B" ~ parse_number(DAMAGE_PROPERTY) * 10^9,
  is.na(DAMAGE_PROPERTY) ~ 0))
```

```{r}
masterData |> 
  filter(DAMAGE_PROPERTY > 0) |>
  ggplot(aes(x = factor(PHASE), y = log(DAMAGE_PROPERTY))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red")
```

This graph displays the distribution of the logarithm-transformed property damage for each ENSO phase, with the red dots representing the average damage. This visualization aimed to explore whether the ENSO phase influences tornado-related property damage.

The box plots for each phase are very similar, showing only minor differences in interquartile range and number of outliers. The El Nino and Neutral phases have near identical distributions and averages, with the Neutral phase having more extreme high-end outliers. La Nina, however, was noticeable in having an increased median and average, along with an upward-shifted range. While property damage is not the best measurement in tornado strength, since other factors also play a role, the data indicates that ENSO has little to no impact on tornado damage.

```{r}
tornado_ranks <- masterData |> group_by(PHASE, TOR_F_SCALE) |> summarise(count = n()) |>  ungroup()
tornado_ranks |> 
  filter(TOR_F_SCALE != "EFU") |>
  ggplot(aes(x = TOR_F_SCALE, y = count, fill = PHASE)) +
  geom_bar(stat = "identity") +
  facet_wrap(~PHASE)
```

Another indicator of a tornado's strength is its EF rating, which is based on observed damage such as uprooted trees, destroyed roofs, etc. This bar chart displays the distribution of tornado ratings since 2007 for each ENSO phase. Tornadoes with a rating of EFU (Enhanced Fujita Unknown) have been excluded because there was insufficient information to determine their rating.

All three ENSO phases display right-skewed distributions, with EF0 and EF1 tornadoes being the most common. The distributions' shapes are very similar; however, La Nina stands out with relatively stronger tornadoes compared to El Nino and Neutral. The shapes of El Nino and Neutral are nearly identical, but El Nino is more likely to have EF0 and EF1 tornadoes. Unlike property damage, where ENSO had little to no influence, the EF ratings show that La Nina is likelier to have stronger tornadoes. In contrast, El Nino is likelier to have weaker ones.

```{r}
masterData |> ggplot(aes(x = TOR_LENGTH, y = TOR_WIDTH, color = PHASE)) +
  geom_point()+
  facet_wrap(~PHASE)

```

The final graph is a scatter plot displaying tornado width on the y-axis and path length on the x-axis, with each point representing an individual tornado observation. A pattern emerges across all three phases where path length shows little variability while tornado width shows greater variability. This likely reflects that certain local conditions have a greater impact on tornado size, causing greater variability. In contrast, tornado lifespan relies on certain conditions being met, regardless of strength or size, leading to less variability. Since the distribution shape for all three phases is nearly identical, it suggests that ENSO does not have a meaningful impact on tornado size and path length.

------------------------------------------------------------------------

## Conclusion

In conclusion, the analysis has revealed an evident relationship between ENSO phases and tornado formation. Initially, I was aware that ENSO impacted the jet stream, which could impact tornado activity, but I did not know to what extent. I came in expecting that the impact on the jet stream would greatly influence every aspect of tornado formation, but that was not the case. Although there was a clear indication that El Nino had an effect through noticeable changes in size and region of activity compared to the Neutral phase, La Nina did not have as many noticeable changes, indicating an influence compared to the Neutral phase. One surprising observation was how little size and path length were affected, as I thought a change in the jet stream would have a larger influence than it did. But other factors, such as the activity window, were more evident between the ENSO phases. Therefore, this analysis has revealed that ENSO influences tornado formation and activity, but many other factors also influence tornadoes.

Further research and analysis will be needed to explore how much of an impact ENSO has on tornado formation and activity. Firstly, I would balance the number of observations for each phase to help clarify the individual impact, as one limitation to this analysis was that the Neutral phase dominated the data set. Second, I would categorize the full year as the dominant ENSO phase instead of working with the individual months, as it created more noise in the dataset and made the analysis more complex. Finally, it would be valuable to look into what other phenomenon interacts with ENSO in tornado formation to better isolate its specific impact.
